How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-25 talk/dyla2007/talk.tex
More shuffling.

\begin{frame}
  \frametitle{Auto-generating VMs}
  \begin{itemize}
  \item
    high-level source: early design decisions not necessary
  \item
    we need a custom \emph{translation toolchain} to compile the interpreter
    to a full VM
  \item
    many aspects of the final VM are orthogonal from the interpreter source:
    they are inserted during translation
  \item
    ``monads'', with more ad-hoc control
  \end{itemize}
  \pause
  \begin{block}{
    Examples}
    \begin{itemize}
    \item
      Garbage Collection strategy
    \item
      Threading models (e.g.\ coroutines with CPS...)
    \item
      non-trivial translation aspect: auto-generating a dynamic compiler from
      the interpreter % XXX maybe we really want to write more about this
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Good Points of the Approach}
  \begin{itemize}
  \item
    dynamic languages can be implemented in a high level language
  \item
    doing this separates high-level from low-level concerns
  \item
    generating various VMs from a single source interpreter, which is easy to
    write
  \item
    lots of flexibility at all levels:
    \begin{itemize}
    \item
      when writing the interpreter
    \item
      also possible to adapt the translation toolchain if necessary
    \item
      this allows breaking of abstraction barriers, when necessary (e.g. tagged
      integers)
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Open Issues / Drawbacks / Further Work}
  \begin{itemize}
  \item
    writing the translation toolchain in the first place takes lots of effort
    (but it can be reused)
  \item
    writing a good GC is still necessary. but: can we maybe reuse existing good
    GCs (e.g. from the Jikes RVM)?
  \item
    many abstraction layers
  \item
    dynamic compiler generation seems to work, but needs to be improved
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion / Meta-Points}
  \begin{itemize}
  \item
    high-level languages are suitable to implement dynamic languages
  \item
    doing so has many benefits
  \item
    VM's shouldn't be written by hand
  \item
    PyPy's concrete approach is not so important
  \item
    diversity is good
  \item
    let's write more meta-programming tool-chains!
  \end{itemize}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Auto-generating VMs}
  \begin{itemize}
  \item
    high-level source: early design decisions not necessary
  \item
    we need a custom \emph{translation toolchain} to compile the interpreter
    to a full VM
  \item
    many aspects of the final VM are orthogonal from the interpreter source:
    they are inserted during translation
  \item
    translation aspect $\cong$ monads, with more ad-hoc control
  \end{itemize}
  \pause
  \begin{block}{
    Examples}
    \begin{itemize}
    \item
      Garbage Collection strategy
    \item
      Threading models (e.g.\ coroutines with CPS...)
    \item
      non-trivial translation aspect: auto-generating a dynamic compiler from
      the interpreter
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Good Points of the Approach}
  {\bf Simplicity:}

  \begin{itemize}
  \item
    dynamic languages can be implemented in a high level language
  \item
    separation of concerns from low-level details
  \item
    a potential single-source-fits-all interpreter --
    less duplication of efforts
  \item
    runs everywhere with the same semantics --
    no outdated implementations, no ties to any standard platform
  \end{itemize}
  \pause
  \begin{block}{PyPy}
     arguably the most readable Python implementation so far
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Good Points of the Approach}
  {\bf Flexibility} at all levels:

  \begin{itemize}
  \item
    when writing the interpreter (high-level languages rule!)
  \item
    when adapting the translation toolchain as necessary
  \item
    to break abstraction barriers when necessary
  \end{itemize}
  \pause
  \begin{block}{
    Example}
    \begin{itemize}
    \item
      boxed integer objects, represented as tagged pointers
    \item
      manual system-level RPython code
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Good Points of the Approach}
  {\bf Performance:}

  \begin{itemize}
  \item
    ``reasonable'' performance
  \item
    can generate a dynamic compiler from the interpreter
    (work in progress, 60x faster on very simple Python code)
  \end{itemize}
  \pause
  \begin{block}{
    JIT compiler generator}
    \begin{itemize}
    \item
      almost orthogonal from the interpreter source - applicable to many
      languages, follows language evolution ``for free''
    \item
      based on Partial Evaluation
    \item
      benefits from a high-level interpreter and a tweakable translation
      toolchain
    \item
      generating a dynamic compiler is easier than generating a static one!
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Open Issues / Drawbacks / Further Work}
  \begin{itemize}
  \item
    writing the translation toolchain in the first place takes lots of effort
    (but it can be reused)
  \item
    writing a good GC is still necessary. but: maybe we can reuse existing good
    GCs (e.g.\ from the Jikes RVM)?
  \item
    conceptually simple approach but many abstraction layers
  \item
    dynamic compiler generation seems to work, but needs more efforts.
    also: can we layer it on top of the JIT of a general purpose OO VM?
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion / Meta-Points}
  \begin{itemize}
  \item
    high-level languages are suitable to implement dynamic languages
  \item
    doing so has many benefits
  \item
    VMs shouldn't be written by hand
  \item
    PyPy's concrete approach is not so important
  \item
    diversity is good
  \item
    let's write more meta-programming toolchains!
  \end{itemize}
\end{frame}
